\paperTitle

### 2014-05-18 doc/icml2014/abstract.tex
minor edits

Recent work on the method of moments enable
  consistent parameter estimation, but
  only for certain types of latent-variable models. %, avoiding local optima issues,
  %but are applicable only to certain types of graphical models.
On the other hand, pure likelihood objectives, though more universally applicable,
  are difficult to optimize.
In this work, we show that using the method of moments in conjunction
  with composite marginal likelihood
  yields consistent parameter estimates for a much broader
  class of discrete directed and undirected graphical models,
  including loopy graphs with high treewidth.
%In particular, our algorithm applies to any graphical model wherein each
  %hidden variable has at least three conditionally independent observed
  %variables.
  %that sidestep local optima issues for a general class of
  %directed and undirected graphical models.
%Parameter estimation in latent-variable models based on EM
  %is prone to local optima.
  %due to the non-convexity of the marginal log-likelihood. 
%In this paper, we extend recent advances in the method of moments 
  %to provide consistent parameter estimates
  %that sidestep local optima issues for a general class of
  %directed and undirected graphical models.
%In particular, our algorithm applies to any graphical model wherein each
  %hidden variable has at least three conditionally independent observed
  %variables.
%This family includes loopy graphs with high treewidth.
Specifically, we first use tensor factorization to reveal information
about the hidden variables.  This renders the otherwise non-convex
likelihood objective convex, and EM can then be used to obtain the global optimum.
%We use the method of moments to tie each hidden to some observed
  %variables, and show that the rest of the problem can be cast as convex
  %optimization of the likelihood.
%Our approach gracefully extends to models outside our class by
  %incorporating the partial information via posterior regulraization.
\end{abstract}

>>>>

Recent work on the method of moments enable
  consistent parameter estimation, but
  only for certain types of latent-variable models.
On the other hand, pure likelihood objectives, though more universally applicable,
  are difficult to optimize.
In this work, we show that using the method of moments in conjunction
  with composite likelihood
  yields consistent parameter estimates for a much broader
  class of discrete directed and undirected graphical models,
  including loopy graphs with high treewidth.
Specifically, we use tensor factorization to reveal information
about the hidden variables.
This allows us to construct convex likelihoods which can be globally optimized
to recover the parameters.
\end{abstract} 
%This renders the otherwise non-convex
%likelihood objective convex, and EM can then be used to obtain the global optimum.

### 2014-05-18 doc/icml2014/discussion.tex
minor edits

% Observable operator
Another prominent line of work in the method of moments community has
  focused on recovering {\em observable operator
  representations} \citep{jaeger2000observable,hsu09spectral,bailly2010spectral,balle12automata}.
  These methods allow prediction of new observations, but do not
  recover the actual parameters of the model, making them difficult to use
  in conjunction with likelihood-based models. % which are driven by parameters.
\citet{song2011spectral} proposed an algorithm to learn observable
  operator representations for latent tree graphical models, like the
  one in \figureref{examples-tree}, assuming the graph is bottlenecked. 
Their approach is similar to our first step of learning conditional moments,
  but they only consider trees.
\citet{parikh12spectral} extended this approach to general graphical
  models which are bottlenecked using a latent junction tree representation. 
  Consequently, the size of the observable representations is exponential in
  the treewidth.
  In contrast, our algorithm only constructs moments of the order of size of the cliques
  (and sub-neighborhoods for pseudolikelihood), which can be much smaller.
For example, an $n\times n$ grid model has a treewidth of $n$, but each
  clique is of size at most $2$ and the degree is at most $4$.
%In some sense, an observable operator re-formulation of our algorithm
  %subsumes this work.

% Identifiability
An interesting direction is to examine the necessity of the bottleneck
property.  Certainly, more than two views is necessary to ensure
identifiability \cite{kruskal77three}, but requiring \emph{each} hidden variable to be
a bottleneck is stronger than we would like.  We hope that by judiciously
leveraging likelihood-based methods in conjunction with the method of moments,
we can generate new hybrid techniques for estimating
an even richer class of latent-variable models.

>>>>

% Observable operator
Another prominent line of work in the method of moments community has
  focused on recovering {\em observable operator
  representations} \citep{jaeger2000observable,hsu09spectral,bailly2010spectral,balle12automata}.
  These methods allow prediction of new observations, but do not
  recover the actual parameters of the model, making them difficult to use
  in conjunction with likelihood-based models. % which are driven by parameters.
\citet{song2011spectral} proposed an algorithm to learn observable
  operator representations for latent tree graphical models, like the
  one in \figureref{examples-tree}, assuming the graph is bottlenecked. 
Their approach is similar to our first step of learning conditional moments,
  but they only consider trees.
\citet{parikh12spectral} extended this approach to general graphical
  models which are bottlenecked using a latent junction tree representation. 
  Consequently, the size of the observable representations is exponential in
  the treewidth.
  In contrast, our algorithm only constructs moments of the order of size of the cliques
  (and sub-neighborhoods for pseudolikelihood), which can be much smaller.
%In some sense, an observable operator re-formulation of our algorithm
  %subsumes this work.

% Identifiability
An interesting direction is to examine the necessity of the bottleneck
property.  Certainly, three views is in general needed to ensure
identifiability \cite{kruskal77three}, but requiring \emph{each} hidden variable to be
a bottleneck is stronger than what we would like.  We hope that by judiciously
leveraging likelihood-based methods in conjunction with the method of moments,
we can generate new hybrid techniques for estimating
even richer classes of latent-variable models.

### 2014-05-18 doc/icml2014/undirected.tex
minor edits

%\renewcommand{\algorithmicrequire}{\textbf{Input:}}
%\renewcommand{\algorithmicensure}{\textbf{Output:}}
\begin{algorithm}
  \caption{\LearnParameters}
  \label{algo:undirected}
  \begin{algorithmic}
    \REQUIRE Conditional moments $\mOpp{v}{i} = \Pr(x_v \mid h_i)$
             and hidden marginals $Z_S = \Pr(\bh_S)$.
    \ENSURE Parameters $\theta$.
    \IF{$\sG$ is directed} \STATE Normalize each $\Pr(a,\Pa(a))$. 
    \ELSIF{$\sG$ is undirected with low treewidth}
      \STATE Compute expected features $\{\mu_\sC\}_{\sC \in \sG}$ (\equationref{logLinearFeatures}).
      \STATE Optimize full likelihood (\equationref{logLinearSupervised}).
    \ELSIF{$\sG$ is undirected with high treewidth}
      \STATE Compute expected features $\{\mu_S\}_{S \in \sB}$ (\equationref{neighborhoodFeatures}).
      \STATE Optimize pseudolikelihood (\equationref{pseudolikelihood}).
    \ENDIF
    \end{algorithmic}
\end{algorithm}

@@

While we now have a complete algorithm for estimating directed and undirected
models, optimizing the full likelihood (\equationref{logLinearSupervised})
can still be computationally intractable for undirected models with high treewidth
due to the intractability of the log-partition function $A(\theta)$.
One can employ various variational approximations of $A(\theta)$ \cite{wainwright08varinf},
but these generally lead to inconsistent estimates of $\theta$.
We thus turn to an older idea of pseudolikelihood \citep{besag75pseudo}.
The pseudolikelihood objective is a sum over the log-probability of each
variable $a$ given its neighbors $\sN(a)$:
\begin{align}
  \label{eqn:pseudolikelihood}
  L_\text{pseudo}(\theta) \eqdef \E_{(\bx,\bh) \sim p_{\theta^*}}\left[\sum_{a \in X \cup H} \log p_\theta(a \mid \sN(a))\right].
\end{align}
In the fully-supervised setting, it is well-known that pseudolikelihood
provides consistent estimates which are computationally efficient
but less statistically efficiency.\footnote{
Coincidentally, this is the same high-level motivation for using method of
moments in the first place.}

@@

A key observation is that for each sub-neighborhood $B \in \sB(a)$, each $\{a\} \cup B$ is bidependent:
conditioning on $a$ does not introduce new independencies within $B$ by construction of $\sB(a)$,
and conditioning on any $b \in B$ does not either since every other $b' \in B \backslash \{b\}$
is connected to $a$.
Assuming $\sG$ is bottlenecked, by \lemmaref{bottleneck-views}
we have that $\{a\} \cup B$ has exclusive views.
Hence, we can recover $\Pr(a,B)$ for each $a$ and $B \in \sB(a)$.
Based on conditional independence of the sub-neighborhoods $B$ given $a$,
we have that $\Pr(a, \sN(a)) = \Pr(a) \prod_{B \in \sB(a)} \Pr(B \mid a)$.
This allows us to estimate the expected features $\mu_{a,\sN(a)}$ and use these quantities
in the optimization of the pseudolikelihood objective.

@@

Note that our pseudolikelihood-based approach
does depend exponentially on the size of the sub-neighborhoods,
which could be larger than the largest clique size.
Therefore,
for this approach to be effective,
each node essentially should have low degree or locally exhibit a lot of
conditional independence.
%While this method is only viable for graphs with small sub-neighboof low
%degree, i.e.\@ when $|\sN(a)|$ is small,
On the positive side, we can handle graphical models with high treewidth;
neither sample nor computational complexity depends on the treewidth.
For example, in the grid model, the treewidth is linear in the size of the grid,
but it is the degree (only 4 in this case) that governs the computational complexity.

>>>>

%\renewcommand{\algorithmicrequire}{\textbf{Input:}}
%\renewcommand{\algorithmicensure}{\textbf{Output:}}
\begin{algorithm}
  \caption{\LearnParameters}
  \label{algo:undirected}
  \begin{algorithmic}
    \REQUIRE Conditional moments $\mOpp{v}{i} = \Pr(x_v \mid h_i)$
             and hidden marginals $Z_S = \Pr(\bh_S)$.
    \ENSURE Parameters $\theta$.
    \IF{$\sG$ is directed} \STATE Normalize $\Pr(a,\Pa(a))$ for $a \in \bx \cup \bh$.
    \ELSIF{$\sG$ is undirected with low treewidth}
      \STATE Compute features $\mu_\sC$ for $\sC \in \sG$ (\equationref{logLinearFeatures}).
      \STATE Optimize full likelihood (\equationref{logLinearSupervised}).
    \ELSIF{$\sG$ is undirected with high treewidth}
      %\STATE Compute expected features $\mu_{\{a\} \cup B}$ for $a \in \bh, B \in \sB(a)$ (\equationref{neighborhoodFeatures}).
      \STATE Compute features $\mu_{\{a\} \cup \sN(a)}$ for $a \in \bh$ (\equationref{neighborhoodFeatures}).
      \STATE Optimize pseudolikelihood (\equationref{pseudolikelihood}).
    \ENDIF
    \end{algorithmic}
\end{algorithm}

@@

While we now have a complete algorithm for estimating directed and undirected
models, optimizing the full likelihood (\equationref{logLinearSupervised})
can still be computationally intractable for undirected models with high treewidth
due to the intractability of the log-partition function $A(\theta)$.
One can employ various variational approximations of $A(\theta)$ \cite{wainwright08varinf},
but these generally lead to inconsistent estimates of $\theta$.
We thus turn to an older idea of pseudolikelihood \citep{besag75pseudo}.
The pseudolikelihood objective is a sum over the log-probability of each
variable $a$ given its neighbors $\sN(a)$:
\begin{align}
  \label{eqn:pseudolikelihood}
  L_\text{pseudo}(\theta) \eqdef \E_{(\bx,\bh) \sim p_{\theta^*}}\left[\sum_{a \in \bx \cup \bh} \log p_\theta(a \mid \sN(a))\right].
\end{align}
In the fully-supervised setting, it is well-known that pseudolikelihood
provides consistent estimates which are computationally efficient
but less statistically efficiency.\footnote{
Coincidentally, this is the same high-level motivation for using method of
moments in the first place.}

@@

A key observation is that for each sub-neighborhood $B \in \sB(a)$, each $\{a\} \cup B$ is bidependent:
conditioning on $a$ does not introduce new independencies within $B$ by construction of $\sB(a)$,
and conditioning on any $b \in B$ does not either since every other $b' \in B \backslash \{b\}$
is connected to $a$.
Assuming $\sG$ is bottlenecked, by \lemmaref{bottleneck-views}
we have that $\{a\} \cup B$ has exclusive views.
Hence, we can recover $\Pr(a,B)$ for each $a$ and $B \in \sB(a)$.
Based on conditional independence of the sub-neighborhoods $B$ given $a$,
we have that $\Pr(a, \sN(a)) = \Pr(a) \prod_{B \in \sB(a)} \Pr(B \mid a)$.
This allows us to compute the expected features $\mu_{a,\sN(a)}$ and use them
in the optimization of the pseudolikelihood objective.

@@

Note that our pseudolikelihood-based approach
does depend exponentially on the size of the sub-neighborhoods,
which could be larger than the largest clique size.
Therefore,
for this approach to be effective,
each node essentially should have low degree or locally exhibit a lot of
conditional independence.
%While this method is only viable for graphs with small sub-neighboof low
%degree, i.e.\@ when $|\sN(a)|$ is small,
On the positive side, we can handle graphical models with high treewidth;
neither sample nor computational complexity necessarily depends on the treewidth.
%For example, for an $n \times n$ grid model, the treewidth is $n$,
%but it is the degree (only 4 in this case) that governs the computational complexity.
For example, an $n\times n$ grid model has a treewidth of $n$, but each
  clique is of size at most $2$ and the degree is at most $4$.
